---
title: "Retail Transactions:  Monthly Report"
author: "Rohit Shivthare"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r seed}
set.seed(41)
```

```{r libraries}
library(data.table)
library(DT)
library(lmtest)
library(dplyr)
```

```{r functions}

# 1. Define a function to round numeric values to specified digits
round.numerics <- function(x, digits){
  if(is.numeric(x)){
    x <- round(x = x, digits = digits)
  }
  return(x)
}

# 2. Define a function to get the top N viewed products for each category
get_top_n_viewed <- function(category_data, n) {
  return(head(category_data[order(-views_count)], n))
}

# 3. Define a function to build a linear regression model and return the spending increase amount with the input extra income
lm_IncreaseSpend_WithIncome <- function(extra_income){
 
  # Filter transactions for the selected month 
  transactions_for_period <- transactions_dt[date >= start_date & date <= end_date]
  
  # Calculate total spending for each customer ID
  total_spending <- transactions_for_period[,.(total_spending = sum(price * quantity)), by = customer_id]
  
  # Merge total spending with customer data (age, gender, income, region)
  customer_spend_data <- left_join(customers, total_spending, by = "customer_id")
  
  # Build a linear regression model
  model <- lm(total_spending ~ income + age + gender + region, data = customer_spend_data)
  
  # Extract the coefficient for income from the summary
  summary_result <- summary(model)
  income_coefficient <- summary_result$coefficients["income", "Estimate"]
  
  # Calculate the increase in monthly average spending associated with an extra income
  increase_in_spending <- income_coefficient * extra_income
  
  # Round the result to two decimal places
  increase_in_spending <- round.numerics(increase_in_spending, 2)
  
  # Return the increase in monthly average spending 
  return(increase_in_spending)
}

# 4. Define a function to create age groups in the dataset
create_age_groups <- function() {

# Merge product_count with customers and create an 'age_group' column
customer_with_products <- merge(product_count, customers[, c("age", "customer_id")], by = "customer_id", all = TRUE)
customer_with_products[, age_group := cut(age, breaks = c(18, 34, 49, 64, Inf), labels = c('18-34', '35-49', '50-64', '65+'))]

# Filter out rows with NA in 'views_count' and 'age_group'
customer_with_products <- customer_with_products[!is.na(views_count) & !is.na(age_group)]

return(customer_with_products)
}

# 5. Define a function to display Top 10 most viewed products for every unique category
display_Top_10 <- function(unique_categories, category_data){
  
for (unique_category in unique_categories) {
  cat("Category:", unique_category, "\n")
  top10_category_data <- category_data[unique_category == top_10_viewed_by_category$category, ][1:10]
  print(top10_category_data)
  cat("\n")
}
}
```

```{r load_data}

setwd("/Users/rohit/Downloads")

customers = read.csv("../Data/customers.csv")

products = read.csv("../Data/products.csv")

if (start_date == "2020-01-01") {
    transactions = read.csv("../Data/transactions -- January 2020.csv")
    views = read.csv("../Data/views -- January 2020.csv")
} else {
    transactions = read.csv("../Data/transactions -- February 2020.csv")
    views = read.csv("../Data/views -- February 2020.csv")
}

```

```{r explore_data, eval = FALSE}
str(customers)
str(products)
str(transactions)
str(views)
```

```{r clean_data}

# Split the 'time' column into 'date' and 'time' columns
transactions$date <- as.Date(substring(transactions$time, 1, 10))
transactions$time <- format(as.POSIXct(transactions$time, format = "%Y-%m-%dT%H:%M:%OS", tz = "UTC"), format = "%H:%M:%S")

# Split the 'time' column into 'date' and 'time' columns
views$date <- as.Date(substring(views$time, 1, 10))
views$time <- format(as.POSIXct(views$time, format = "%Y-%m-%dT%H:%M:%OS", tz = "UTC"), format = "%H:%M:%S")
```

```{r constants}

# Create a data.table for 'views'
views_dt <- as.data.table(views)

# Create a data.table for 'transactions' 
transactions_dt <- as.data.table(transactions)

## Create a data.table for 'products'
products_dt <- as.data.table(products)

# Create a data.table for 'customers'
customers_dt <- as.data.table(customers)

# Filter views for the selected month
views_for_period <- views_dt[date >= start_date & date <= end_date]

# Filter transactions for the selected month
transactions_for_period <- transactions_dt[date >= start_date & date <= end_date]
```


## Specific Questions {.tabset}

This part of the report will be directed to product managers throughout the client's company.  The idea is to give them the useful information they need to act on the specific questions they posed.  Plan your communication accordingly.

For this report, make sure to use all of the data that was provided to you for the month.

### Q1

#### Question

During the first 7 days of the month of January, what were the 10 most viewed products?  Show the results in a table with the product's identifier, category, and count of the number of views.

#### Answer

```{r part2_q1}

# Filter views for the first 7 days of selected month
first_week_views <- views_dt[date >= first_week_start & date <= first_week_end]

# Count the number of views for each product
product_views_count <- first_week_views[, .(views_count = .N), by = product_id]

# Select the top 10 most viewed products
top_10_viewed_products <- product_views_count[order(-views_count)][1:10]

# Join with the 'products' data to get product information
top_10_viewed_products <- merge(top_10_viewed_products, products, by = "product_id")

# Sort the top 10 viewed products in descending order
top_10_viewed_products <- top_10_viewed_products[order(-views_count)]

# Display the results using DT
datatable(top_10_viewed_products, 
          options = list(pageLength = 10, scrollX = TRUE))

```

### Q2

#### Question

During the whole month, what were the 10 most viewed products for each category?  Show the results in separate tables by category.  Include only the product's identifier and the count of the number of views.


#### Answer {.tabset}

```{r part2_q2_overall_calculations}

# Join 'views_for_period' with 'products' to include category information
views_with_category <- merge(views_for_period, products, by = "product_id")

# Count the number of views for each product in each category
category_views_count <- views_with_category[, .(views_count = .N), by = .(category, product_id)]

# Get the top 10 most viewed products for each category
top_10_viewed_by_category <- category_views_count[, get_top_n_viewed(.SD, 10), by = category]

# Display the results in separate tables by category
unique_categories <- unique(top_10_viewed_by_category$category)

```

```{All category Top 10 most viewed products, eval = FALSE}
display_Top_10(unique_categories, top_10_viewed_by_category)
```

##### Shirt

```{r part2_q2_shirt}
cat("Category:", "Shirt", "\n")
category_data <- top_10_viewed_by_category[top_10_viewed_by_category$category == "shirt",][1:10]
print(category_data)
```

##### Pants

```{r part2_q2_pants}
cat("Category:", "Pants", "\n")
category_data <- top_10_viewed_by_category[top_10_viewed_by_category$category == "pants",][1:10]
print(category_data)
```

##### Coat

```{r part2_q2_coat}
cat("Category:", "Coat", "\n")
category_data <- top_10_viewed_by_category[top_10_viewed_by_category$category == "coat",][1:10]
print(category_data)
```

##### Shoes

```{r part2_q2_shoes}
cat("Category:", "Shoes", "\n")
category_data <- top_10_viewed_by_category[top_10_viewed_by_category$category == "shoes",][1:10]
print(category_data)
```

##### Hat

```{r part2_q2_hat}
cat("Category:", "Hat", "\n")
category_data <- top_10_viewed_by_category[top_10_viewed_by_category$category == "hat",][1:10]
print(category_data)
```


### Q3

#### Question

What was the total revenue for each category during the month?  Show the results in a single table sorted in decreasing order.

#### Answer

```{r part2_q3}

# Perform an inner join between 'transactions_dt' and 'products_dt' on 'product_id'
merged_dt <- transactions_for_period[products_dt, on = "product_id"]

# Calculate revenue for each transaction
merged_dt[, revenue := price * quantity]

# Sum revenue by category and sort in decreasing order
category_revenue <- merged_dt[, .(total_revenue = sum(revenue)), by = category]

# Format total_revenue with commas
category_revenue[, total_revenue := format(total_revenue, big.mark = ",", scientific = FALSE)]

# Sort in descending order
category_revenue <- category_revenue[order(-total_revenue)]

# Print the results
print(category_revenue)
```

### Q4

#### Question

Among customers with at least one transaction, show the average, median, and standard deviation of monthly spending per customer.

#### Answer

```{r part2_q4}

## Calculate monthly spending per customer
#customer_stats <- transactions_for_period[, .(average = mean(price * quantity), 
#                                           median = median(price * quantity), 
#                                           stddev = if (.N > 1) sd(price * quantity) else NA_real_), 
#                                       by = customer_id]
#
## Print the results
#print(customer_stats)

## Calculate total spending for each customer
#customer_total_spending <- transactions_for_period[, .(
#  Avg = mean(price * quantity, na.rm = TRUE),
#  Median = median(price * quantity, na.rm = TRUE),
#  StdDev = sd(price * quantity, na.rm = TRUE)
#), by = customer_id]
#
## Calculate the average, median, and standard deviation of spending for all customers
#average_spending <- mean(customer_total_spending$Avg, na.rm = TRUE)
#median_spending <- median(customer_total_spending$Median, na.rm = TRUE)
#std_deviation_spending <- sd(customer_total_spending$StdDev, na.rm = TRUE)
#
## Round the results to two decimal places
#average_spending <- round(average_spending, 2)
#median_spending <- round(median_spending, 2)
#std_deviation_spending <- round(std_deviation_spending, 2)
#
## Print the results
#cat("The average spending is: $", average_spending, "\n")
#cat("The median spending is: $", median_spending, "\n")
#cat("The standard deviation is: $", std_deviation_spending, "\n")


# Calculate total spending for each customer and count the number of transactions
customer_spending_count <- transactions_for_period[, .(spending = sum(price * quantity), count = .N), by = customer_id]

# Filter customers with at least one transaction
customers_with_transactions <- customer_spending_count[count >= 1]

# Calculate the average, median, and standard deviation of spending
result <- customers_with_transactions[, .(
  Average = round(mean(spending), 2),
  Median = round(median(spending), 2),
  Standard_Deviation = round(sd(spending), 2)
)]

# Print the results
cat("The average spending is: $", result$Average, "\n")
cat("The median spending is: $", result$Median, "\n")
cat("The standard deviation is: $", result$Standard_Deviation, "\n")
```

### Q5

#### Question

What percent of the monthly revenue was attributed to each category of gender?  Show the amount of revenue and the percentage for each group.

#### Answer

```{r part2_q5}

## Perform inner joins to combine transactions_for_period, 'products', 'customers'
#merged_dt <- transactions_for_period[products_dt, on = "product_id"]
#merged_dt <- merged_dt[customers_dt, on = "customer_id"]
#
## Calculate revenue for each transaction
#merged_dt[, revenue := price * quantity]
#
## Sum revenue by gender and category
#gender_category_revenue <- merged_dt[, .(total_revenue = sum(revenue)), by = .(gender, category)]
#
## Filter out rows where 'category' is NA
#gender_category_revenue <- gender_category_revenue[!is.na(category)]
#
## Calculate the total revenue for the selected month
#total_monthly_revenue <- sum(gender_category_revenue$total_revenue, na.rm = TRUE)
#
## Calculate the percentage of revenue for each group
#gender_category_revenue[, percentage := (total_revenue / total_monthly_revenue) * 100]
#
## Sort revenue by gender
#setorder(gender_category_revenue, gender)
#
## Format total_revenue with commas
#gender_category_revenue[, total_revenue := format(total_revenue, big.mark = ",", scientific = FALSE)]
#
## Round the percentage column to 2 decimal places
#gender_category_revenue[, percentage := round.numerics(percentage, 2)]
#
## Print the results
#print(gender_category_revenue)

# Perform left joins to combine transactions_for_period, 'products', 'customers'
merged_dt <- transactions_for_period[products_dt, on = "product_id", nomatch = 0L]
merged_dt <- merged_dt[customers_dt, on = "customer_id", nomatch = 0L]

# Calculate revenue for each transaction
merged_dt[, revenue := price * quantity]

# Sum revenue by gender and category
gender_category_revenue <- merged_dt[, .(total_revenue = sum(revenue)), by = .(gender)]

# Calculate the total revenue for the selected month
total_monthly_revenue <- sum(gender_category_revenue$total_revenue, na.rm = TRUE)

# Calculate the percentage of revenue for each group
gender_category_revenue[, percentage := (total_revenue / total_monthly_revenue) * 100]

# Round the percentage column to 2 decimal places
gender_category_revenue[, percentage := round(percentage, 2)]

# Filter rows for 'male', 'female', and 'other' genders
desired_genders <- c('M', 'F', 'Other')
gender_category_revenue <- gender_category_revenue[gender %in% desired_genders]

# Format total_revenue with commas
gender_category_revenue[, total_revenue := format(total_revenue, big.mark = ",", scientific = FALSE)]

# Round the percentage column to 2 decimal places
gender_category_revenue[, percentage := round.numerics(percentage, 1)]

# Print the results
print(gender_category_revenue)
```

### Q6

#### Question

Using linear regression, what is the increase in monthly average spending associated with an extra ten thousand dollars of income?  Make sure your model also includes age, gender, and region as inputs to adjust for.

#### Answer

```{r part2_q6}

increase_in_spending = lm_IncreaseSpend_WithIncome(10000)
increase_in_spending

```


### Q7

#### Question

Among customers who viewed at least 1 product, how many had at least one purchase during the month?  Show the total number of customers with a view, the total who made a purchase, and the percentage of customers with a view who made a purchase.

#### Answer

```{r part2_q7}

# Filter views for customers who viewed at least 1 product during selected month
viewed_customers <- unique(views_for_period$customer_id)

# Filter transactions for customers who made at least 1 purchase during selected month
purchased_customers <- unique(transactions_for_period$customer_id)

# Calculate the total number of customers with a view and customers who made a purchase
total_customers_with_view <- length(viewed_customers)
total_customers_with_purchase <- length(purchased_customers)

# Calculate the percentage of customers with a view who made a purchase
percentage_with_purchase <- (total_customers_with_purchase / total_customers_with_view) * 100

# Create a data.table to display the results
result_dt <- data.table(
  Total_Customers_With_View = total_customers_with_view,
  Total_Customers_With_Purchase = total_customers_with_purchase,
  Percentage_With_Purchase = percentage_with_purchase
)

# Format the total_customers_with_view and total_customers_with_purchase with commas
result_dt[, `:=`(
  Total_Customers_With_View = format(Total_Customers_With_View, big.mark = ",", scientific = FALSE),
  Total_Customers_With_Purchase = format(Total_Customers_With_Purchase, big.mark = ",", scientific = FALSE)
)]

# Round the Percentage_With_Purchase column to 2 decimal places
result_dt[, Percentage_With_Purchase := round.numerics(Percentage_With_Purchase, 2)]

# Print the results
print(result_dt)
```

### Q8

#### Question

Now let's look at the viewing habits in different age groups, including 18-34, 35-49, 50-64, and 65+.  Within each group, what were the mean, median, and standard deviation for the number of unique products viewed per customer?

#### Answer

```{r part2_q8}

customer_with_products <- create_age_groups()

# Calculate mean, median, and standard deviation of 'views_count' by 'age_group'
result <- customer_with_products[, .(
  mean = round(mean(views_count), 2),
  median = round(median(views_count), 2),
  std = round(sd(views_count), 2)
), by = age_group]

# Print the results
print(result)
```


### Q9 

#### Question

What is the correlation between a user's total page views and total spending?  Look only at customers with at least 1 view for the month.  For customers without a transaction, include their spending as zero.

#### Answer

```{r part2_q9}

# Find customers who viewed at least one product in the selected month
viewed_customers <- unique(views_for_period$customer_id)

# Calculate total spending for each customer
total_spending <- transactions_for_period %>%
  filter(customer_id %in% viewed_customers) %>%
  group_by(customer_id) %>%
  summarize(total_spending = sum(price * quantity))

# Calculate total page views for each customer
total_page_views <- views_for_period %>%
  filter(customer_id %in% viewed_customers) %>%
  group_by(customer_id) %>%
  summarize(total_page_views = n())

# Combine total page views and total spending into a single data frame
customer_data <- merge(total_page_views, total_spending, by = "customer_id", all = TRUE)

# Replace missing spending values with 0
customer_data[is.na(customer_data$total_spending), "total_spending"] <- 0

# Calculate the correlation between total page views and total spending
correlation <- cor(customer_data$total_page_views, customer_data$total_spending, use = "complete.obs")

# Round the correlation to two decimal places
correlation <- round.numerics(correlation, 2)

# Print the correlation
correlation
```

### Q10

#### Question

Which customer(s) purchased the largest number of coats during the month?  In the event of a tie, include all of the users who reached this value, listed in increasing sorted order.  Show their identifiers and total volume of coats purchased during the month.

#### Answer

```{r part2_q10}

# Merge the 'products' table to check for the "coat" product
transactions_for_period <- transactions_for_period[products_dt, on = "product_id"]

# Calculate the total volume of coats purchased by each customer
coats_purchased <- transactions_for_period[category == "coat", .(total_coats = sum(quantity)), by = customer_id]

# Find the largest number of coats purchased
max_coats <- max(coats_purchased$total_coats)

# Filter customers who purchased the largest number of coats
top_coat_customers <- coats_purchased[total_coats == max_coats, .(customer_id, total_coats)]

# Sort the top customers by customer_id in increasing order
top_coat_customers <- top_coat_customers[order(customer_id)]

# Print the result
print(top_coat_customers)
```

